### 2021-12-11 documents/sections/proto_carib.tex
more finetuning

Interestingly, all six featured innovative first person markers on \gl{s_a_} verbs.
All other (complete) extensions%
\footnote{As an honorable mention, when \ikpeng replaced third person \setone with \settwo prefixes, \obj{a} \qu{to be} and \obj{ke} \qu{to say} retained \obj{n-} \parencite[12]{matter2019arara}.
	However, the spread of the innovative markers had an entirely different pathway from the extensions discussed here, spreading from subordinate to main clauses, rather than from verbs to verbs.} either occurred with other person values and/or targeted \gl{s_p_} verbs.
Illustrative examples for complete extensions are shown in \cref{tab:completed}: the extension of \gl{1+2}\gl{s_a_} \obj{s(ɨ)-} (< \rc{kɨt-}) to \gl{s_p_} verbs in \apalai \pcref{tab:apalai}, of \gl{2}\gl{s_a_} \obj{m(ɨ)-} in to \gl{s_p_} verbs in \panare \pcref{tab:panare}, and the extension of the entire \gl{s_a_} set to \gl{s_p_} verbs in \waimiri \pcref{tab:waimiri}. %, as well as the extension of \rc{k-} to \gl{1}\gl{s_p_} and of \rc{tɨt-} to \gl{1+2}\gl{s_p_} in \PWai \pcref{fig:par_ext}.
The starkly different behavior of \gl{s_a_} and \gl{s_p_} verbs regarding extensions points to the split-\gl{s} system playing a role, so its main properties will be discussed in \cref{sec:split}.
It will also be made clear how the \gl{s_a_}/\gl{s_p_} distinction can be lost for a single person, or how \gl{s_p_} verbs can take on \gl{s_a_} markers with apparent semantic impunity.

@@

\subsection{The Cariban split-\gl{s} system}
\label{sec:split}
As seen in \cref{sec:pc_person}, the distinction between \gl{s_a_} and \gl{s_p_} verbs was instantiated by inflection classes within the \PC \setone person paradigm, but this was not the only inflectional difference:
%\begin{inlinelist}
%	\item presence vs absence of the \gl{s_a_} marker \rc{w-}
%	\item absence vs presence of the second person prefix \rc{ə(j)-} in imperatives
%	\item presence vs absence of a derivational detransitivizing prefix
%\end{inlinelist}.
%
%todo CONTINUE HERE
\textcite[208]{meira2000split} Many languages show an \gl{s_a_} class marker in deverbalized forms, originating in \PC \rc{w-}.\footnote{See \textcite[227]{meira2000split}, who identifies reflexes of this morpheme as having \dbqu{no purpose other than being \qu{class markers}, without any obvious semantic or functional load}.}
With \gl{s_a_} verbs, \rc{w-} occurred immediately between the possessive prefixes and the verb stem, while \gl{s_p_} verbs took the bare prefixes \pcref{tab:participles,tab:nominalizations}.
%Reflexes of \rc{w-} in languages from different branches are illustrated in \cref{tab:participles,tab:nominalizations} for participles and nominalizations.%
%
%\input{floats/participles}%
%
%\input{floats/nominalizations}%
%
%\input{floats/imperatives}%
%
The distinction between \gl{s_a_} and \gl{s_p_} was also reflected in imperatives, where the latter took the \gl{2}\gl{s_p_} prefix \rc{ə(j)-} while the former were unprefixed .
%This is illustrated with reflexes in various modern languages in \pcref{tab:imperatives}.
%Both the \gl{s_a_} marker \rc{w-} and the prefix pattern in imperatives have been lost in some languages, but are reconstructible to \PC.
%However, there was one further property uniting most \gl{s_a_} verbs, not based on inflectional morphology.

In the modern instantiations of split-\gl{s}, mismatches between the semantics of verbs and their \gl{s_a_} or \gl{s_p_} membership are common, exemplified with \kalina data in \exref{kar2}.
%
%\gl{a}- or \gl{p}-oriented inflectional morphology are common, exemplified with \kalina data in \exref[kar1]{kar2}.
%
%\pex<kar1> \kalina
%\a<kar-60>
%\begingl
%\gla mi-kupi-ja//
%\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
%\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
%\endgl
%\a<kar-62>
%\begingl
%\gla a-kupi-ja//
%\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
%\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
%\endgl
%\xe
%%
%In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
%The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
%The intransitive verbs in \exref{kar2} show the same person markers, but the choice of marker depends on the verb rather than semantics.
%
\pex<kar2> \kalina
\a<kar-59>
\begingl
\gla sipi tɨnka-rɨ m-ekema-non hen//
\glb net pull-\gl{nmlz} \gl{2}-be.afraid-\gl{prs}.\gl{uncert} eh?//
\glft \qu{You're afraid to pull up the net, aren't you? } \parencite[][253]{courtz2008carib}//
\endgl
\a<kar-61>
\begingl
\glpreamble aya:woiya //
\gla aj-awomɨ-ja//
\glb \gl{2}-get.up-\gl{prs}//
\glft \qu{You are getting up.} \parencite[][167]{hoff1968carib}//
\endgl
\xe
%
In \exref{kar2.kar-59}, the \gl{s_a_} verb \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker, while the \gl{s_p_} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.
In both cases, the prefix does not appear to contribute to the semantics of the predicate, since there are clear mismatches:
\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker is clearly volitional.
\textcite{meira2000split} investigates a corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs in any of the languages.

Rather, the reason for a verb to take \gl{a}- or \gl{p}-oriented prefixes is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
	Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
%\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
%\begin{inlinelist}
%	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
%	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
%	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
%	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
%	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
%\end{inlinelist}.
He notes that this leads to an inflectional split not based in meaning, but rather morphology:

@@

Regarding the form of this marker, \textcite[505--512]{meira2010origin} reconstruct two distinct prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although on verbs their reflexes have been merged into a single morpheme in modern languages.
Modern reflexes of \detrz show a range of meanings summarizable as \dbqu{detransitive}, illustrated with \trio \gl{s_a_} verbs in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
	\\
	\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
	\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
	\\
	\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
	\qu{wash} & & \obj{əi-suka} & (reflexive)\\
	\\
	\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
	\qu{break (\gl{tr})} & & & (anticausative)\\
	\\
	\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
	\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe
%
The morphological variation featured in \qu{to abandon each other} and \qu{to wash self} is due to the collapse between the two \PC prefixes:
\obj{e-} is a reflex of the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.
In the next section, it will become evident that many of the verbs not affected by person marker extensions belong to the small group of \gl{s_a_} verbs without a reflex of \detrz.
%As mentioned, there are a few \gl{s_a_} verbs that are not derived from transitive verbs, and do not have a reflex of \detrz, either.
%\textcite[221]{meira2000split} characterizes these as very old and exceptional, and counts between 5 and 7 verbs, depending on the language.
%\textcite{gildea2007greenberg} identify 7 such underived \gl{s_a_} verbs as reconstructible to \PC, shown in \cref{tab:underived2007}.
%Many of these verbs will be discussed in \cref{sec:verbs}.

>>>>

Interestingly, all six featured innovative first person markers on \gl{s_a_} verbs.
All other (complete) extensions%
\footnote{As an honorable mention, when \ikpeng replaced third person \setone with \settwo prefixes, \obj{a} \qu{to be} and \obj{ke} \qu{to say} retained \obj{n-} \parencite[12]{matter2019arara}.
	However, the spread of the innovative markers had an entierly different dynamic, spreading from subordinate to main clauses.} either occurred with other person values and/or targeted \gl{s_p_} verbs.
Illustrative examples for complete extensions are shown in \cref{tab:completed}: the extension of \gl{1+2}\gl{s_a_} \obj{s(ɨ)-} (< \rc{kɨt-}) to \gl{s_p_} verbs in \apalai \pcref{tab:apalai}, of \gl{2}\gl{s_a_} \obj{m(ɨ)-} in to \gl{s_p_} verbs in \panare \pcref{tab:panare}, and the extension of the entire \gl{s_a_} set to \gl{s_p_} verbs in \waimiri \pcref{tab:waimiri}. %, as well as the extension of \rc{k-} to \gl{1}\gl{s_p_} and of \rc{tɨt-} to \gl{1+2}\gl{s_p_} in \PWai \pcref{fig:par_ext}.
The starkly different behavior of \gl{s_a_} and \gl{s_p_} verbs regarding extensions points to the split-\gl{s} system playing a role, so its main properties will be discussed in \cref{sec:split}.
It will also be made clear how the \gl{s_a_}/\gl{s_p_} distinction can be lost for a single person, or how \gl{s_p_} verbs can take on \gl{s_a_} markers with apparent semantic impunity.

@@

\subsection{The Cariban split-\gl{s} system}
\label{sec:split}
As seen in \cref{sec:pc_person}, the split between \gl{s_a_} and \gl{s_p_} verbs was instantiated by inflection classes within the \PC \setone person paradigm, but this was not the only inflectional criterion:
In deverbalized forms, \gl{s_a_} verbs took a class marker \rc{w-}, while \gl{s_p_} verbs lacked that prefix \parencites[89, 141--142]{gildea1998}[208]{meira2000split}.
The distinction between \gl{s_a_} and \gl{s_p_} was also reflected in imperatives, where the latter took the \gl{2}\gl{s_p_} prefix \rc{ə(j)-} while the former were unprefixed \parencites[89]{gildea1998}[208]{meira2000split}.

%\parencite[208]{meira2000split} Many languages show reflexes \gl{s_a_}  in deverbalized forms, originating in .
%With \gl{s_a_} verbs, \rc{w-} occurred immediately between the possessive prefixes and the verb stem, while \gl{s_p_} verbs took the bare prefixes \pcref{tab:participles,tab:nominalizations}.
%Reflexes of \rc{w-} in languages from different branches are illustrated in \cref{tab:participles,tab:nominalizations} for participles and nominalizations.%
%
%\input{floats/participles}%
%
%\input{floats/nominalizations}%
%
%\input{floats/imperatives}%
%
%The distinction between \gl{s_a_} and \gl{s_p_} was also reflected in imperatives, where the latter took the \gl{2}\gl{s_p_} prefix \rc{ə(j)-} while the former were unprefixed .
%This is illustrated with reflexes in various modern languages in \pcref{tab:imperatives}.
%Both the \gl{s_a_} marker \rc{w-} and the prefix pattern in imperatives have been lost in some languages, but are reconstructible to \PC.
%However, there was one further property uniting most \gl{s_a_} verbs, not based on inflectional morphology.

In modern instantiations of Cariban split-\gl{s}, mismatches between the semantics of verbs and their \gl{s_a_} or \gl{s_p_} status are common, exemplified with \kalina data in \exref{kar2}.
%
%\gl{a}- or \gl{p}-oriented inflectional morphology are common, exemplified with \kalina data in \exref[kar1]{kar2}.
%
%\pex<kar1> \kalina
%\a<kar-60>
%\begingl
%\gla mi-kupi-ja//
%\glb \gl{2}>\gl{3}-bathe-\gl{prs}//
%\glft \qu{You bathe him/her.} \parencite[][160]{hoff1968carib}//
%\endgl
%\a<kar-62>
%\begingl
%\gla a-kupi-ja//
%\glb \gl{3}>\gl{2}-bathe-\gl{prs}//
%\glft \qu{S/he bathes you.} \parencite[][63]{yamada2011evidentiality}//
%\endgl
%\xe
%%
%In \exref{kar1}, the choice between the second person \gl{a}- and \gl{p}-oriented markers \obj{mi-} and \obj{a-} depends on the scenario:
%The transitive verb \obj{kupi} \qu{to bathe} takes \obj{mi-} in \gl{2}>\gl{3} scenarios \exref{kar1.kar-60}, but \obj{a-} in \gl{3}>\gl{2} scenarios \exref{kar1.kar-62}.
%The intransitive verbs in \exref{kar2} show the same person markers, but the choice of marker depends on the verb rather than semantics.
%
\pex<kar2> \kalina
\a<kar-59>
\begingl
\gla sipi tɨnka-rɨ m-ekema-non hen//
\glb net pull-\gl{nmlz} \gl{2}\gl{s_a_}-be.afraid-\gl{prs}.\gl{uncert} eh?//
\glft \qu{You're afraid to pull up the net, aren't you? } \parencite[][253]{courtz2008carib}//
\endgl
\a<kar-61>
\begingl
\glpreamble aya:woiya //
\gla aj-awomɨ-ja//
\glb \gl{2}\gl{s_p_}-get.up-\gl{prs}//
\glft \qu{You are getting up.} \parencite[][167]{hoff1968carib}//
\endgl
\xe
%
The \gl{s_a_} verb \obj{ekema} \qu{to be afraid} takes an \gl{a}-oriented marker \exref{kar2.kar-59}, while the \gl{s_p_} verb \obj{awomɨ} \qu{to get up} takes a \gl{p}-oriented marker \exref{kar2.kar-61}.
In both cases, the prefix does not appear to contribute to the semantics of the predicate, since there are clear mismatches:
\qu{to be afraid} with an \dbqu{agentive} marker can hardly be considered a volitional act, while  \qu{to get up} with a \dbqu{patientive} marker is clearly volitional.
\textcite{meira2000split} investigates a corpus of intransitive verbs from \trio, \kalina, \apalai, and \wayana, and categorizes them by applying different criteria commonly encountered in split-\gl{s} systems.
He shows that neither (non\-)activities, % \parencite[210]{meira2000split}, 
(non-)agency, % \citeyearpar[212]{meira2000split}, 
(in-)animacy, % \citeyearpar[213]{meira2000split},
nor Aktionsart % \citeyearpar[215]{meira2000split} 
satisfactorily predict the class membership of intransitive verbs in any of the languages.

Rather, the reason for a verb to take \gl{a}- or \gl{p}-oriented prefixes is (at least diachronically) a morphological one.
\textcite[217--221]{meira2000split} demonstrates that those intransitive verbs which (etymologically) have a derivational detransitivizing prefix are treated as \gl{s_a_} verbs, while essentially all others are \gl{s_p_} verbs:
\begin{quotebox}{\parencite[201]{meira2000split}}
	Almost all verbs in the \gl{s_a_} class are detransitivized forms of transitive verbs, either synchronically (with still exisiting transitive sources) or diachronically (with reconstructible but no longer existing transitive sources)
\end{quotebox}
%\textcite[221--223]{meira2000split} also argues that the detransitivizing prefixes are indeed deriving \gl{s_a_} verbs, rather than being inflectional in nature:
%\begin{inlinelist}
%	\item there are a few underived \gl{s_a_} verbs, with no detransitivizing prefix
%	\item \gl{s_a_} verbs can develop irregular semantics compared to their transitive counterparts
%	\item it is unpredictable whether the \gl{a} or \gl{p} argument of the underlying transitive verb becomes the \gl{s} of the derived \gl{s_a_} verb
%	\item some originally derived \gl{s_a_} verbs have lost their transitive counterparts
%	\item \dbqu{basic} concepts are expressed as derivations of more complex concepts, like \qu{to dance (\gl{s_a_})} from \qu{to dance with (\gl{tr})}
%\end{inlinelist}.
He notes that this leads to an inflectional split not based in meaning, but rather morphology:

@@

Regarding the form of this marker, \textcite[505--512]{meira2010origin} reconstruct two distinct prefixes for \PC: reciprocal \rc{əte-} and reflexive \rc{e-}, although their reflexes on verbs have been merged into a single morpheme in modern languages.
Reflexes of \detrz show a range of meanings summarizable as \dbqu{detransitive}, illustrated with \trio \gl{s_a_} verbs in \exref{triodetrz}.

\ex<triodetrz> \trio \parencites[218--219]{meira2000split}[128, 256]{triomeira1999}\\
\begin{tabular}[t]{@{}llll@{}}
	\\
	\obj{nonta}  & → & \obj{e-nonta}, & \qu{abandon each other}\\
	\qu{abandon} & & \obj{əi-nonta} &  (reciprocal) \\
	\\
	\obj{suka} & → & \obj{e-suka}, & \qu{wash self}\\
	\qu{wash} & & \obj{əi-suka} & (reflexive)\\
	\\
	\obj{pahka} & → & \obj{e-pahka} & \qu{break (\gl{intr})}\\
	\qu{break (\gl{tr})} & & & (anticausative)\\
	\\
	\obj{puunəpɨ} & → & \obj{əh-puunəpɨ}, & \qu{think, meditate}\\
	\qu{think about} & & \obj{əi-puunəpɨ} & (antipassive)\\
\end{tabular}
\xe
%
The morphological variation in \qu{to abandon each other} and \qu{to wash self} is due to the collapse of the two \PC prefixes:
\obj{e-} comes from the reflexive prefix \rc{e-}, while the form \obj{əi-} originates in reciprocal \rc{əte-}.
However, both can occur with either meaning -- at least for these two verbs.
In the next section, it will become apparent that many of the verbs not affected by person marker extensions belong to the small group of \gl{s_a_} verbs without a reflex of \detrz.
%As mentioned, there are a few \gl{s_a_} verbs that are not derived from transitive verbs, and do not have a reflex of \detrz, either.
%\textcite[221]{meira2000split} characterizes these as very old and exceptional, and counts between 5 and 7 verbs, depending on the language.
%\textcite{gildea2007greenberg} identify 7 such underived \gl{s_a_} verbs as reconstructible to \PC, shown in \cref{tab:underived2007}.
%Many of these verbs will be discussed in \cref{sec:verbs}.
